---
title: "The effects of floral resource removal on 
plant-pollinator network structure"
author: "Justin Bain"
date: "March 21, 2019"
output:
  word_document: default
  html_document:
    df_print: paged
---

# 1. Introduction 

Understanding the dynamics and stability of ecological communities is a central question in ecology. With increasing anthropogenic and global climate change pressures, understanding how communities respond to environmental disturbances has become even more urgent. Pollination is a critical ecosystem service that may be threatened in many natural systems by various global change pressures including habitat loss and climate change. Most flowering plants rely on pollinators for reproduction, while pollinators utilize floral resources for food, making plant-pollinator interactions a delicate relationship in which population declines to either group can negatively affect the other (Willmer 2011). Recent studies have shown that global change pressures can have severe impacts on plants and pollinators. For example, in a community in Illinois that has lost substantial amounts of natural habitat, 50% of bee species have been locally extirpated over the last 120 years, reducing the quantity and quality of pollination services over time (Burkle et al. 2013).

Environmental disturbances can have substantial impacts on community dynamics. A shifting floral resource landscape can introduce new species interactions, eliminate existing interactions or otherwise influence existing interactions indirectly. Ecological networks (e.g., plant-pollinator interaction networks) are powerful tools for understanding the structure of complex sets of interactions among all species in a community (Bascompte and Jordano 2014). Interaction networks are composed of sets of species where the links depict interactions among species. When all links among all species are considered, the emergent structure of the network can be quantified using several descriptive metrics that describe different properties of the interactions (Bascompte and Jordano 2014). Network metrics (e.g., connectance, nestedness, and specialization) can provide useful information regarding the stability and resilience of the network to various perturbations and disturbances (Thebault and Fontaine 2010; Burkle et al. 2013).

My project is addressing how environmental disturbance may impact plant-pollinator interactions by presenting one week of data from an experiment I co-lead in 2017 where we experimentally removed a dominant floral resource from an intact subalpine plant community and explored its effects on plant-pollinator interaction network structure. My analysis contains four main components: (i) calculate the network metrics, (ii) create bipartite networks for control, removal, and overall networks, (iii) plot the network metrics using ggplot2, and (iv) conduct paired t-tests to determine whether any differences in network metrics are statistically significant. 


## Goal of Project
To understand how the removal of a dominant floral resource, *Helianthella quinquenervis*, impacts plant-pollinator network structure.


## Question: How does the removal of *Helianthella* influence plant-pollinator network structure? 

## Hypotheses

* **Null**: Network metrics will not differ between control and removal treatments. 

* **Alternative**: Network metrics will differ between control and removal treatments.

* **Prediction 1**: Removing *Helianthella* will lead to higher network specialization (higher H2’) as pollinators will floral food resources available to them.

* **Prediction 2**: Removing *Helianthella* will increase nestedness due to specialists on Helianthella needing to shift their resource use or disappear from the removal networks altogether. Alternatively, if pollinators do not shift their resource-use, then the removal networks will likely be less nested than control networks.

# 2. Materials and Methods 

## Data Source

https://github.com/j-a-bain/HQnetwork 

This dataset was collected by Justin A. Bain, Rachel G. Dickson, and Paul J. CaraDonna at the Rocky Mountain Biological Laboratory in Gothic, CO from July - August 2017. The dataset contains records of plant-pollinator interactions in plots that contain Helianthella quinquenervis and plots where H. quinquenervis was removed. There were 20 paired plots (10 removal and 10 control plots). Observations were paired (by J.A.B and R.G.D) and each plot was observed for 15 minutes. An interaction was recorded when a floral visitor contacted the reproductive structures of the plant. All weather data were recorded prior to the start of each observation set and, except for cloud cover, all weather data were recorded using a Kestrel 5000. 

## Load packages
```{r message=FALSE, warning=FALSE}
library(bipartite) #for network analyses
library(tidyverse) # loads ggplot2, dplyr (manipulate data), and tidyr (convert long to wide)
library(data.table) # manipulating data frames
library(reshape2) # for melt() function
library(plyr) # data summaries
library(igraph) # for fancy network plots
```

## Read in data file
```{r message=FALSE, warning=FALSE}
#read in data file
hq.data<-read_csv("https://www.dropbox.com/s/4yqi9eu2pbgijhw/Final_Project_Data.csv?dl=1")

glimpse(hq.data)
```

## Clean up data
* I first select only relevant columns for data analysis.
* Then I filter observations that cannot be used for analysis.
* Finally, I check my dataset using glimpse().
```{r message=FALSE, warning=FALSE}
# filter/subset data
hq.data1 <- hq.data %>% 
  select(date,week,pair_number,plot,treatment,
         plant, pollinator, num_total_visits, 
         num_unique_observations, keep_remove) %>% 
  filter(!pollinator == "bombus_spp",  
        !keep_remove == 'remove') 
glimpse(hq.data1)
```


## Calculate Network Metrics
* First I split the data into networks for each week and treatment combination. There will be 10 control networks and 10 removal treatment networks.
* In order to create bipartite networks and calculate network metrics using ‘bipartite’, data frames must be turned into a matrix format. To turn each plot into a matrix, I run xtabs() on nets.split to turn each list into a matrix. Using function(x) allows xtabs to run on all lists at once.
*	Next, I use lapply and function(x) to calculate all network metrics at once on all matrices. lapply() applies whatever is in the parentheses to all lists. The function(x) calculates metrics across all matrices. Metrics are calculated using networklevel() from the bipartite package.
* Then I organize the network metric results output by binding each list into a single data frame using do.call(rbind,…). I use melt() to convert the data frame from wide to long format.
* The next step is to make sure the column names are appropriate. I first split the first column into multiple columns for week, treatment, and pair number and assign new names to those columns. 
* Finally, I combine the new column names with the results data frame using cbind() and assign new column names. Finally, save the results using write.csv.

```{r}
#turn data into networks for each week x treatment combination
nets.split <- split(hq.data1, list(hq.data1$week, hq.data1$treatment, hq.data1$pair_number), drop=TRUE) 

# calculate network metrics:
# 1. use lapply to turn each list into a matrix of plant-pollinator interactions

# reformat all lists so they are a matrix (i.e., "table")
networkz <- lapply(nets.split, function(x) xtabs(num_total_visits ~ pollinator + plant, droplevels(x))) 

# 2. calculate metrics on all matrices
nets.all.vals.output<-lapply(networkz, function(x) 
networklevel(x, index=c("connectance", "weighted connectance", "NODF", "weighted NODF", "links per species", "H2")))

# 3. organize network metric results output
net.vals.all <- do.call("rbind", nets.all.vals.output)
# The melt function converts from a wide format to a long format, making the data "tidy". 
net.vals.all.melt <- melt(net.vals.all)
# add more appropriate names. 
colnames(net.vals.all.melt)<-c("net.id", "metric", "value")

#split the net.id column in two to create a week v. cumulative (pooled) column
net.id_split_names <- data.frame(do.call('rbind', strsplit(as.character(net.vals.all.melt$net.id),".",fixed=TRUE))) 

#combine the split names with the resutls data frame
net.vals.all.ready <- cbind(net.vals.all.melt, net.id_split_names)

# apply new names to the data frame
colnames(net.vals.all.ready)<-c("net.id", "metric", "value", "week", "treatment", "pair_number")

# 4. save results as a new csv file.

# write network metrics to csv
#  write.csv(net.vals.all.ready,       
# "hq.data.network.metrics.subset_12MAR2019.csv")
```

# 3. Results

## Create bipartite networks
* Goal: Make networks using the bipartite package.
* In order to create separate control and removal treatment networks, each treatment must be filtered out from the overall data set.
* I again transform the data into a matrix by running xtabs() with the response first followed by interacting groups (i.e., plants and pollinators).
* Once data is in a matrix format, I create the bipartite networks by using plotweb() on each matrix, which creates a network output. 

```{r}
# 1. Control Network
hq.data1.control <- hq.data1 %>% filter(treatment == "control")

control.nets <- xtabs(hq.data1.control$num_total_visits ~ hq.data1.control$pollinator + hq.data1.control$plant)
control.net <- matrix(control.nets, nrow= 11, byrow=TRUE)

# bipartite package networks
plotweb(control.nets)

# 2. Removal Network
hq.data1.removal <- hq.data1 %>% filter(treatment == "removal")
removal.nets <- xtabs(hq.data1.removal$num_total_visits ~ hq.data1.removal$pollinator + hq.data1.removal$plant)
removal.net <- matrix(removal.nets, nrow= 8, byrow=TRUE)

# bipartite package networks
plotweb(removal.nets)

# 3. Overall Network
total.nets <- xtabs(hq.data1$num_total_visits ~ hq.data1$pollinator + hq.data1$plant)
total.net <- matrix(total.nets, nrow=12, byrow=TRUE)

# bipartite package networks
plotweb(total.nets)
```

## Create igraph networks
* Goal: To make more visually appealing bipartite networks using igraph.
* The layout is set to 'layout.circle' to present a circular network but other formats are available. 
* V = vertex attributes (i.e., nodes); E = Edge attributes (i.e, lines)
* First convert the network to an incidence matrix. Then add other features to make the graph look nicer, like changing the node colors and sizes, and line thicknesses.
```{r fig.height=10, fig.width=10}
# Overall Visitation Network
i_net = graph_from_incidence_matrix(total.net, weight=T) # first convert the network to an incidence matrix.
V(i_net)$frame.color = "white"
V(i_net)$label = NA # removes number labels from nodes
polcol = rep("grey35",dim(total.net)[2]) # colors for pollinators
plantcol = rep("honeydew3",dim(total.net)[1]) # colors for plants
clrs = rbind(as.matrix(plantcol),as.matrix(polcol)) # combine the colors as a matrix
V(i_net)$color = clrs # assign the colors to the i_net
V(i_net)$size = 13 # changes the size of the nodes
E(i_net)$width = E(i_net)$weight*0.1 # changes the widths of the lines
plot(i_net,layout=layout.circle) # plot the network
mtext("Overall Visitation Network", side=3, adj=0.8, cex=2.5) # assign text labels.

# Control Network
i_net = graph_from_incidence_matrix(control.net, weight=T)
V(i_net)$frame.color = "white"
V(i_net)$label = NA
polcol = rep("grey35",dim(control.net)[2])
plantcol = rep("honeydew3",dim(control.net)[1])
clrs = rbind(as.matrix(plantcol),as.matrix(polcol))
V(i_net)$color = clrs
V(i_net)$size = 13 
E(i_net)$width = E(i_net)$weight*0.1 
plot(i_net,layout=layout.circle)
mtext("Control Treatment Network", side=3, adj=0.8, cex=2.5)

# Removal Network
i_net = graph_from_incidence_matrix(removal.net, weight=T)
V(i_net)$frame.color = "white"
V(i_net)$label = NA
polcol = rep("grey35",dim(removal.net)[2])
plantcol = rep("honeydew3",dim(removal.net)[1])
clrs = rbind(as.matrix(plantcol),as.matrix(polcol))
V(i_net)$color = clrs
V(i_net)$size = 13
E(i_net)$width = E(i_net)$weight*0.1
plot(i_net,layout=layout.circle)
mtext("Removal Treatment Network", side=3, adj=0.8, cex=2.5)
```

## Metric Figures

* I use the network metrics results file to make figures comparing network metrics for each treatment.
* First, I load the CSV file and remove unneeded columns. 
* Next, I create summaries for each metric using summarise() and ddply() to turn the results into a data frame. I calculate descriptive statistics like the mean, median, minimum, and maximum, and the standard error and 95% confidence interval for each metric. 
* I then use ggplot2() to create plots of the mean (using geom_point()) for each metric in both treatments with 95% confidence intervals as error bars ( using geom_errorbar()).


### Load metric csv file
```{r message=FALSE, warning=FALSE}
# Read in Metric Results CSV
hq.net.results<-read_csv("https://www.dropbox.com/s/gob4ozqtaz20l3f/hq_removal_plot_network_metrics_PBC470_13MAR2019.csv?dl=1")
names(hq.net.results)

# remove the first column (X1) since it's no longer needed
hq.net.results['X1'] <- NULL
hq.net.results
glimpse(hq.net.results)
```

### Metric Plots
* Definitions for each metric are provided in the code below.
```{r}
# weighted connectance: how many interactions were observed out of all possible interactions

# filter data for weighted connectance metric only
nets.w.c <- net.vals.all.ready %>% filter(metric == "weighted connectance")

# create a summary for weighted connectance 
w.connectance.data.sum <- ddply(nets.w.c, c("treatment"), 
                              summarise,
               N    = length(value),
               mean = mean(value),
               median = median(value),
               min = min(value),
               max = max(value),
               sd   = sd(value),
               se   = sd / sqrt(N),
               ci   = qt(0.975, N - 1)* se)
w.connectance.data.sum

# create plot of weighted connectance
ggplot(w.connectance.data.sum, aes(x = treatment, 
                           y = mean)) +
  geom_errorbar(aes(ymin = mean-ci, ymax = mean+ci), 
                colour="#666666", 
                width=.1) +
  geom_point(alpha = 1.0, size = 2.5,
               colour = c("gold3","darkolivegreen4")) +
  ylim(0, 0.3) +
  theme_classic() +
  ylab("Mean Weighted Connectance") +
  ggsave("w.connectance.2019.png", 
         width = 2.5, 
         height = 2.5, 
         dpi = 1200)
```
```{r}
# Linkage density: The average number of links per species

# filter data for links per species metric only
nets.links <- net.vals.all.ready %>% filter(metric == "links per species")

# create a summary for links per species 
links.data.sum <- ddply(nets.links, c("treatment"), 
                              summarise,
               N    = length(value),
               mean = mean(value),
               median = median(value),
               min = min(value),
               max = max(value),
               sd   = sd(value),
               se   = sd / sqrt(N),
               ci   = qt(0.975, N - 1)* se)
links.data.sum

# create plot of links per species
ggplot(links.data.sum, aes(x = treatment, 
                           y = mean)) +
  geom_errorbar(aes(ymin = mean-ci, ymax = mean+ci), 
                colour="#666666", 
                width=.1) +
  geom_point(alpha = 1.0, size = 2.5,
               colour = c("gold3","darkolivegreen4")) +
  theme_classic() +
  ylim(0,1.1) +
  ylab("Mean Links Per Species") +
  ggsave("links.2019.png", 
         width = 2.5, 
         height = 2.5, 
         dpi = 1200)

```
```{r}
# H2’: Network-level index of specialization.  (higher number = more specialized)

# filter data for H2 metric only
nets.H2 <- net.vals.all.ready %>% filter(metric == "H2")

# create a summary for H2
H2.data.sum <- ddply(nets.H2, c("treatment"), 
                              summarise,
               N    = length(value),
               mean = mean(value),
               median = median(value),
               min = min(value),
               max = max(value),
               sd   = sd(value),
               se   = sd / sqrt(N),
               ci   = qt(0.975, N - 1)* se)
H2.data.sum

# create plot of H2
ggplot(H2.data.sum, aes(x = treatment, 
                           y = mean)) +
  geom_errorbar(aes(ymin = mean-ci, ymax = mean+ci), 
                colour="#666666", 
                width=.1) +
  geom_point(alpha = 1.0, size = 2.5,
               colour = c("gold3","darkolivegreen4")) +
  ylim(0,1) +
  theme_classic() +
  ylab("Mean Specialization (H2)") +
  ggsave("H2.2019.png", 
         width = 2.5, 
         height = 2.5, 
         dpi = 1200)

```
```{r}
# Weighted Nestedness (NODF): Specialist species interact with a subset of the species that generalist species interact with.

# filter data for w.NODF metric only
nets.w.NODF <- net.vals.all.ready %>% filter(metric == "weighted NODF")

# create a summary for w.NODF
w.NODF.data.sum <- ddply(nets.w.NODF, c("treatment"), 
                              summarise,
               N    = length(value),
               mean = mean(value),
               median = median(value),
               min = min(value),
               max = max(value),
               sd   = sd(value),
               se   = sd / sqrt(N),
               ci   = qt(0.975, N - 1)* se)
w.NODF.data.sum

# create plot of w.NODF
ggplot(w.NODF.data.sum, aes(x = treatment, 
                           y = mean)) +
  geom_errorbar(aes(ymin = mean-ci, ymax = mean+ci), 
                colour="#666666", 
                width=.1) +
  geom_point(alpha = 1.0, size = 2.5,
               colour = c("gold3","darkolivegreen4")) +
  ylim(0,30) +
  theme_classic() +
  ylab("Mean Weighted NODF") +
  ggsave("w.NODF.2019.png", 
         width = 2.5, 
         height = 2.5, 
         dpi = 1200)
```

## Paired T-tests
* To determine whether there are statistically significant differences between treatments for each metric, I run paired t-tests for each metric and create a table of the results for easy reference. I use paired t-tests because my plots and observations are paired and using a paired test provides more statistical power than an unpaired t-test.
```{r}
# weighted connectance t-test 
w.connectance.tt <- t.test(nets.w.c$value ~ nets.w.c$treatment, paired=T)
w.connectance.tt

# links per species t-test 
links.tt <- t.test(nets.links$value ~ nets.links$treatment, paired=T)
links.tt

# H2 t-test 
H2.tt <- t.test(nets.H2$value ~ nets.H2$treatment, paired=T)
H2.tt

# weighted NODF t-test 
w.NODF.tt <- t.test(nets.w.NODF$value ~ nets.w.NODF$treatment, paired=T)
w.NODF.tt
```

### Table of t-test results

|     Metric         |   t-value| df|  p-value|
|-------------------:|---------:|--:|--------:|
|w.connectance       |    -10.36|  9|  < 0.001|
|links per species   |      2.45|  9|    0.037|
|specialization (H2')|     -2.37|  9|    0.042|
|w.nestedness (NODF) |      2.52|  9|    0.033|

# 4. Conclusions

The paired t-tests showed that the mean metric values significantly differed between treatments, confirming that the removal of Helianthella had an impact on plant-pollinator interactions and consequentially interaction network structure. Therefore I can reject the null hypothesis that the network metrics would differ between control and removal treatments. I used paired t-tests because the plots were paired, which provides more statistical power to detect differences between groups.

The control networks have higher average nestedness (t = 2.52; p = 0.03) and links per species (t = 2.45; p = 0.04) than the removal networks, which supports my prediction. These results indicate that the control networks may be more stable than the removal networks as both results indicate higher network generalization. A more generalized network is predicted to confer greater ecological stability than a specialized network (Thebault and Fontaine 2010), as it indicates greater pollinator flexibility and more redundancy in interactions, i.e., a generalist pollinator can possibly make up for the loss of a specialist pollinator (Kaiser-Bunbury et al. 2010). Previous research near the Rocky Mountain Biological Laboratory has shown the pollinator community here to display high rates of pollinator foraging flexibility throughout the growing season (CaraDonna et al. 2017), which supports my results here.

The removal networks were more connected (t = -10.4; p < 0.01) and more specialized (t = -2.37; p = 0.04) than the control networks, partially supporting my prediction. A more specialized network indicates the opposite of a generalized network: less redundancy in interactions and less pollinator flexibility, thus indicating less stability in the removal networks. A more connected network is predicted to confer greater network stability because that means there were more realized interactions between plants and pollinators. However, the higher connectance in the removal networks may be a product of the removal networks containing fewer plant species than the control networks (N = 8 plant species vs. N = 11 species in control networks). 

To better understand the impacts of removing Helianthella on plant-pollinator network structure, I will need to analyze the other two weeks from this study. That will provide insight into how plant-pollinator interactions respond to the loss of Helianthella throughout the flowering period of Helianthella. Also, I will explore some species-level metrics, like centrality and specialization metrics, to understand whether individual species vary in their foraging breadth and overall network role between control and removal networks.
My findings illustrate the diversity of responses that can occur with the removal of a dominant floral resource, with implications for our understanding of the stability and dynamics of ecological interactions in response to various disturbances.


# 5. References
Data Source: https://github.com/j-a-bain/HQnetwork
1.	Bascompte, J., and P. Jordano. 2014. Mutualistic networks. Princeton University Press.
2.	Burkle, L. A., J. C. Marlin, and T. M. Knight. 2013. Plant-Pollinator Interactions over 120 Years: Loss of Species, Co-Occurrence, and Function. Science 339:1611-1615.
3.	CaraDonna, P. J., W. K. Petry, R. M. Brennan, J. L. Cunningham, J. L. Bronstein, N. M. Waser, and N. J. Sanders. 2017. Interaction rewiring and the rapid turnover of plant-pollinator networks. Ecology Letters 20:385-394.
4.	Kaiser-Bunbury, C. N., S. Muff, J. Memmott, C. B. Müller, and A. Caflisch. 2010. The robustness of pollination networks to the loss of species and interactions: A quantitative approach incorporating pollinator behaviour. Ecology Letters 13:442–452.
5.	Thebault, E., and C. Fontaine. 2010. Stability of Ecological Communities and the Architecture of Mutualistic and Trophic Networks. Science 329:853-856.
6.	Willmer, P. 2011. Pollination and floral ecology. Princeton University Press.

